%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                         %
% CHAPTER 01:                                             %
% What are quantum dots?                                  %
%                                                         %
% This file is part of a BSc Thesis Project. See the      %
% LICENSE file for more information about licensing.      %
%                                                         %
% Author:     Matteo Seclì <secli.matteo@gmail.com>       %
% A.Y.:       2014/2015                                   %
% URL:        https://github.com/matteosecli/QMC          %
%                                                         %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction}

Everyone has an intuitive idea of what temperature is in everyday life: it is a quantity that tells us how \emph{hot} or \emph{cold} an object is. \\
Temperature is an intensive property of a system, which means that it does not depend on the object's dimensions nor on the quantity of matter contained in it. \\
Temperature is measured via thermometers and, since it is a physical quantity, it must be exressed in terms of physical units. All thermometers are expected to give the same number when used to measure the temperature of the same object, provided that 
the number is expressed always in the same physical unit. This can be possible if the thermometers are accurately calibrated according to some critera that make the calibration process reproducible. 
A roughly good calibration of a thermometer in the so called Celsius scale can be done by assigning the instrument a value of $0^\circ$C (0 degrees Celsius) at the freezing point of water, and $100^\circ$C at the boiling point of water. A more sophisticated and precise way to calibrate a thermometer 
on the Celsius scale consists in assigning the value of $-273.15^\circ$C to the \emph{coldest} temperature admitted according to nowadays' physics, and $0.01^\circ$ C to the triple point of water, that 
is the point in which water coexists in liquid, solid and gaseous form. One can then extract a measure of all the other values of temperature by assuming a linear relation between temperature and the property in the material used as a thermometer,
such as an elastic deformation. \\
Another interesting temperature scale is the one measured in Kelvin (K), called the \emph{absolute temperature}. This scale of temperature is 
defined in a way such that it gives a values of $0$ K at the coldest possible situation admitted by physics, the so called \emph{absolute zero} point: this means that no physical systems can be cooled more than a system 
whose absolute temperature is $0$ K. \par
\vspace{10pt}
The purpose of this thesis is to introduce the concept of \emph{negative absolute temperatures}. This does not constitute a contradiction to what told before because negative absolute temperatures should not be searched below the absolute zero, but rather 
above infinity: negative absolute temperatures are hotter than all the positive ones. The world of negative absolute temperatures is often accompanied by strange phenomena: for example, when a system at negative temperature is put into contact with one at positive temperature, the former gives heat to the latter,
leading the temperature of the former to be lowered and the temperature of the latter to be raised. \par
\vspace{10pt}
In the first chapter of this thesis I will introduce the concept of temperature in a rigorous way according to physical laws, justifying the existence of negative absolute temperatures basing on thermodynamical arguments. \\
A simple system that admits negative temperatures, namely the two level system, will then be presented in chapter \ref{ch:TLS}. \\
In chapter \ref{ch:PandP} I will show that negative absolute temperatures were experimentally observed by Purcell and Pound in 1950. 
After that discovery, some criticisms were moved against the definition of the entropy used to define the absolute scale, putting in doubt the existence of negative temperatures: this will be discussed in chapter \ref{ch:entropy}. \\
Finally some simulations of a systems at negative temperature will be presented in the last chapter.